---
title: "Getting Started with PLS-DA Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with PLS-DA Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Introduction

This vignette demonstrates how to use the `maternal.immune` package to perform Partial Least Squares Discriminant Analysis (PLS-DA) on cytokine data to predict maternal immune status.

## Load the Package

```{r setup}
library(maternal.immune)
library(ggplot2)
```

## Example Data

For this example, we'll create simulated cytokine data with two groups representing different maternal immune status levels.

```{r create-data}
set.seed(42)

# Simulate cytokine measurements for 100 samples
n_samples <- 100
n_cytokines <- 15

# Create outcome variable (maternal immune status)
immune_status <- factor(rep(c("yes", "no"), each = n_samples / 2))

# Simulate cytokine data with some group differences
cytokines <- matrix(rnorm(n_samples * n_cytokines),
                    nrow = n_samples,
                    ncol = n_cytokines)

# Add some discriminatory signal to certain cytokines
# Cytokines 1-5 will have higher values in "yes" group
for (i in 1:5) {
  cytokines[immune_status == "yes", i] <-
    cytokines[immune_status == "yes", i] + rnorm(n_samples/2, mean = 1.5, sd = 0.5)
}

# Cytokines 6-8 will have lower values in "yes" group
for (i in 6:8) {
  cytokines[immune_status == "yes", i] <-
    cytokines[immune_status == "yes", i] - rnorm(n_samples/2, mean = 1, sd = 0.5)
}

# Name the cytokines
colnames(cytokines) <- paste0("IL", 1:n_cytokines)

# Display dimensions
cat("Data dimensions:\n")
cat("  Samples:", nrow(cytokines), "\n")
cat("  Cytokines:", ncol(cytokines), "\n")
cat("  Outcome classes:", table(immune_status), "\n")
```

## Step 1: Validate the Data

Before running the analysis, it's good practice to validate the input data.

```{r validate}
# Validate the data
validate_plsda_data(cytokines, immune_status)
```

## Step 2: Prepare the Data

The `prepare_plsda_data()` function helps preprocess the data by:
- Removing zero-variance predictors
- Handling missing values
- Optionally applying transformations

```{r prepare}
prepared <- prepare_plsda_data(
  X = cytokines,
  Y = immune_status,
  remove_zero_var = TRUE,
  na_action = "fail",  # We have no missing data in this example
  log_transform = FALSE,
  verbose = TRUE
)
```

## Step 3: Run PLS-DA Analysis

Now we can run the PLS-DA model using 2 components.

```{r run-plsda}
result <- run_plsda(
  X = prepared$X,
  Y = prepared$Y,
  ncomp = 2,
  scale = TRUE
)

# Print summary
print(result)
```

## Step 4: Examine VIP Scores

Variable Importance Projection (VIP) scores help identify which cytokines are most important for discriminating between groups. Generally, VIP > 1 indicates an important variable.

```{r vip-scores}
# View top 10 VIP scores
head(result$vip_scores, 10)

# Identify important cytokines (VIP > 1)
important_cytokines <- result$vip_scores[result$vip_scores$vip_score > 1, ]
cat("\nNumber of important cytokines (VIP > 1):", nrow(important_cytokines), "\n")
print(important_cytokines)
```

## Step 5: Visualize VIP Scores

Create a bar plot of the VIP scores to visualize which cytokines are most discriminatory.

```{r plot-vip, fig.width=8, fig.height=6}
# Plot top 15 cytokines
vip_plot <- plot_vip(
  result$vip_scores,
  top_n = 15,
  threshold = 1.0,
  title = "Cytokine Importance for Maternal Immune Status"
)

print(vip_plot)
```

## Step 6: Access the mixOmics Model

The underlying mixOmics model can be accessed for additional analyses.

```{r mixomics-plots, eval=FALSE}
# Score plot (sample plot)
mixOmics::plotIndiv(
  result$model,
  comp = c(1, 2),
  group = result$Y,
  legend = TRUE,
  title = "PLS-DA Sample Plot"
)

# Loading plot (variable plot)
mixOmics::plotLoadings(
  result$model,
  comp = 1,
  title = "Loadings on Component 1"
)
```

## Step 7: Model Performance Assessment

You can assess model performance using cross-validation.

```{r performance, eval=FALSE}
# Perform cross-validation
cv_result <- mixOmics::perf(
  result$model,
  validation = "Mfold",
  folds = 5,
  nrepeat = 10,
  progressBar = FALSE
)

# View performance metrics
plot(cv_result)
```

## Interpretation

### VIP Scores

In this example:
- Cytokines with VIP > 1 are considered important predictors
- The top-ranked cytokines (IL1-IL5 in our simulation) show the strongest discriminatory power
- These cytokines should be prioritized for further biological investigation

### Practical Recommendations

1. **VIP Threshold**: While VIP > 1 is commonly used, you may adjust this threshold based on your specific needs
2. **Component Selection**: Use cross-validation to determine the optimal number of components
3. **Data Scaling**: Scaling is recommended when cytokines are measured on different scales
4. **Sample Size**: Ensure adequate samples per class (minimum 10-15 recommended)

## Summary

This vignette demonstrated:
1. Data preparation and validation
2. Running PLS-DA analysis
3. Calculating and interpreting VIP scores
4. Visualizing important cytokines
5. Accessing the underlying model for advanced analyses

For more information on PLS-DA methodology, see the [mixOmics documentation](http://mixomics.org/).
